cardinalR: Generating Interesting High-Dimensional Data Structures

### 2025-12-12 paper-cardinalR.Rmd
Put clustering stats in a new script (speed up the paper rendering)

```{r load-libraries}
library(cardinalR)
library(tidyverse)
library(kableExtra)
library(geozoo)
library(patchwork)
library(colorspace)
library(crosstalk)
library(fpc)
library(mclust)
```

@@

```{r}
# Extract high-dimensional data
# bad practice!
#data <- five_clusts[, -5]
true_labels <- as.numeric(gsub("cluster", "", five_clusts$cluster))

# Hierarchical clustering
dist_mat <- dist(five_clusts[, -5])
hc_fit <- hclust(dist_mat, method = "ward.D2")
hc_cl <- NULL
for (ncl in 2:10) {
  hcl <- cutree(hc_fit, ncl)
  hc_cl <- cbind(hc_cl, hcl)
}
colnames(hc_cl) <- paste0("cl", 2:10)

x <- cluster.stats(dist_mat,
  hc_cl[,1])[c("within.cluster.ss", "pearsongamma", "dunn2", 
                    "wb.ratio", "ch", "sindex")]
hc_stats <- tibble(method="hc", cl=2) |>
  bind_cols(as_tibble(x))
for (i in 3:10) {
  x <- cluster.stats(dist_mat,
    hc_cl[,i-1])[c("within.cluster.ss", "pearsongamma", "dunn2", 
                    "wb.ratio", "ch", "sindex")]
  hcms <- tibble(method="hc", cl=i) |>
    bind_cols(as_tibble(x))
  hc_stats <- hc_stats |>
    bind_rows(hcms)
}
```

```{r}
# K-means clustering
kmeans_cl <- NULL
for (ncl in 2:10) {
  kcl <- kmeans(five_clusts[, -5], centers = ncl, nstart = 20)$cluster
  kmeans_cl <- cbind(kmeans_cl, kcl)
}
colnames(kmeans_cl) <- paste0("cl", 2:10)

x <- cluster.stats(dist_mat,
  kmeans_cl[,1])[c("within.cluster.ss", "pearsongamma", "dunn2", 
                    "wb.ratio", "ch", "sindex")]
kmeans_stats <- tibble(method="km", cl=2) |>
  bind_cols(as_tibble(x))
for (i in 3:10) {
  x <- cluster.stats(dist_mat,
    kmeans_cl[,i-1])[c("within.cluster.ss", "pearsongamma", "dunn2", 
                    "wb.ratio", "ch", "sindex")]
  kms <- tibble(method="km", cl=i) |>
    bind_cols(as_tibble(x))
  kmeans_stats <- kmeans_stats |>
    bind_rows(kms)
}
```

```{r}
# Model-based clustering
modelbased_cl <- NULL
for (ncl in 2:10) {
  mcl <- Mclust(five_clusts[, -5], G = ncl, modelNames = "EII")$classification
  modelbased_cl <- cbind(modelbased_cl, mcl)
}
colnames(modelbased_cl) <- paste0("cl", 2:10)

x <- cluster.stats(dist_mat,
  modelbased_cl[,1])[c("within.cluster.ss", "pearsongamma", "dunn2", 
                    "wb.ratio", "ch", "sindex")]
modelbased_stats <- tibble(method="mb", cl=2) |>
  bind_cols(as_tibble(x))
for (i in 3:10) {
  x <- cluster.stats(dist_mat,
    modelbased_cl[,i-1])[c("within.cluster.ss", "pearsongamma", "dunn2", 
                    "wb.ratio", "ch", "sindex")]
  mbs <- tibble(method="mb", cl=i) |>
    bind_cols(as_tibble(x))
  modelbased_stats <- modelbased_stats |>
    bind_rows(mbs)
}
```

```{r, label = "fig-cluster-stats", fig.width=10, fig.height=3, out.width="100%", layout="l-body", fig.cap="Comparison of clustering performance metrics (within–between ratio (wb.ratio), Dunn index, Corrected Rand index, and variation of information (VI) across $k$-means, hierarchical, and model-based clustering methods."}

# Examine the cluster stats
all_stats <- bind_rows(hc_stats, kmeans_stats, modelbased_stats)
all_stats |>
  pivot_longer(within.cluster.ss:sindex, names_to = "stat", values_to = "value") |>
  ggplot(aes(x=cl, y=value, colour=method)) +
    geom_line() +
    facet_wrap(~stat, ncol=3, scales="free_y") +
    xlab("Number of clusters") +
    ylab("") +
    scale_color_manual(values = c('#1b9e77','#d95f02','#7570b3')) +
    theme_minimal() +
    theme(panel.border = element_rect(fill = 'transparent'),
          plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
          axis.ticks.x = element_line(),
          axis.ticks.y = element_line(),
          legend.position = "bottom",
          axis.text.x = element_text(size = 7),
          axis.text.y = element_text(size = 7),
          axis.title.x = element_text(size = 7),
          axis.title.y = element_text(size = 7),
          plot.margin = margin(0, 0, 0, 0))
```

Figure \@ref(fig:fig-cluster-stats) shows a selection of cluster metrics for $2-10$ clusters for each of the methods, $k$-means, hierarchical, and model-based. As is typical, the suggestion of the best solution varies between cluster statistics. Although the metrics differ in their preferences, several show consistent support for a $4–5$ cluster solution. The Calinski–Harabasz index increases sharply up to four or five clusters before levelling off, the Dunn and Pearson gamma indices attain local maxima in this range, and both the silhouette width and within-cluster sum of squares show diminishing improvement beyond five clusters. Together, these patterns indicate that $4–5$ clusters provide a good balance of separation and compactness, with $k$-means performing slightly better than the other methods across most metrics.

>>>>

```{r load-libraries}
library(cardinalR)
library(tidyverse)
library(kableExtra)
library(geozoo)
library(patchwork)
library(colorspace)
library(crosstalk)
library(fpc)
library(mclust)
library(htmltools)
```

@@

<!--script/05_gen_cluster_stat.R-->
```{r, label = "fig-cluster-stats", fig.width=10, fig.height=3, out.width="100%", layout="l-body", fig.cap="Cluster validity metrics for solutions with $2–10$ clusters obtained using $k$-means, hierarchical, and model-based clustering. Several indices consistently suggest that $4–5$ clusters provide the best balance of separation and compactness, with $k$-means performing slightly better across metrics."}

# Examine the cluster stats
all_stats <- read_rds("data/five_clusts/cluster_stat.rds")
all_stats |>
  pivot_longer(within.cluster.ss:sindex, names_to = "stat", values_to = "value") |>
  ggplot(aes(x=cl, y=value, colour=method)) +
    geom_line() +
    facet_wrap(~stat, ncol=3, scales="free_y") +
    xlab("Number of clusters") +
    ylab("") +
    scale_color_manual(values = c('#1b9e77','#d95f02','#7570b3')) +
    theme_minimal() +
    theme(panel.border = element_rect(fill = 'transparent'),
          plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
          axis.ticks.x = element_line(),
          axis.ticks.y = element_line(),
          legend.position = "bottom",
          axis.text.x = element_text(size = 7),
          axis.text.y = element_text(size = 7),
          axis.title.x = element_text(size = 7),
          axis.title.y = element_text(size = 7),
          plot.margin = margin(0, 0, 0, 0))
```

Figure \@ref(fig:fig-cluster-stats) shows a selection of cluster metrics for $2-10$ clusters for each of the methods, $k$-means, hierarchical, and model-based. As is typical, the suggestion of the best solution varies between cluster statistics. Although the metrics differ in their preferences, several show consistent support for a $4-5$ cluster solution. The Calinski–Harabasz index increases sharply up to four or five clusters before levelling off, the Dunn and Pearson gamma indices attain local maxima in this range, and both the silhouette width and within-cluster sum of squares show diminishing improvement beyond five clusters. Together, these patterns indicate that $4-5$ clusters provide a good balance of separation and compactness, with $k$-means performing slightly better than the other methods across most metrics.

```{r}
## k-means with  four clusters
kcl4 <- kmeans(five_clusts[, -5], centers = 4, nstart = 20)$cluster

## k-means with  five clusters
kcl5 <- kmeans(five_clusts[, -5], centers = 5, nstart = 20)$cluster
```

```{r highd-data-clusters-algo-html, eval=knitr::is_html_output(), fig.cap="Viewing five synthetic clusters with distinct geometric structures: a helical spiral, a hemisphere, a uniform cube, a cone, and a Gaussian cluster.", fig.alt="Interactive langevitour of five synthetic clusters in $4\\text{-}D$: helical spiral, hemisphere, uniform cube, cone, and Gaussian cluster, each colored distinctly to highlight geometric structure.", fig.pos="!ht", layout = "l-body"}

four_km_langevitour <- langevitour::langevitour(
  five_clusts[, -5],
  group = kcl4,
  levelColors = c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e'),
  width = "400px",
  height = "300px"
)

five_km_langevitour <- langevitour::langevitour(
  five_clusts[, -5],
  group = kcl5,
  levelColors = c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e'),
  width = "400px",
  height = "300px"
)

km_results <- browsable(
  div(
    style = "
      display: flex;
      justify-content: space-between;
      gap: 10px;
    ",
    div(style = "flex: 1;", four_km_langevitour),
    div(style = "flex: 1;", five_km_langevitour)
  )
)

class(km_results) <- c(class(km_results), "htmlwidget")

km_results
```

```{r five-4clusts-projections1}
scaled_data <- scale_data_manual(five_clusts[, -5])

proj_obj1[["cluster"]] <- as.character(kcl4)

five_clusts_proj1_1 <- plot_proj(
  proj_obj = proj_obj1, 
  point_param = c(1.5, 0.2), # size, alpha, color
  plot_limits = c(-1.1, 1), 
  title = "a1", 
  cex = 2, 
  axis_text_size = 4,
  is_color = TRUE) + scale_color_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e')) +
  theme(legend.position = "none")

```{r five-4clusts-projections2}
## Second projection
projection <- cbind(
  c(0.34673,-0.35774,0.66272,0.27298),
  c(0.24023,0.01170,-0.41886,0.72707))

proj_obj2[["cluster"]] <- as.character(kcl4)

five_clusts_proj2_1 <- plot_proj(
  proj_obj = proj_obj2, 
  point_param = c(1.5, 0.2), # size, alpha, color
  plot_limits = c(-1.1, 1.2), 
  title = "a2", 
  cex = 2, 
  axis_text_size = 4,
  is_color = TRUE) + scale_color_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e')) +
  theme(legend.position = "none")
```

```{r five-4clusts-projections3}
## Third projection
projection <- cbind(
  c(0.47388,-0.20984,-0.56560,0.41644),
  c(-0.51124,-0.60184,0.06463,0.36627))

proj_obj3[["cluster"]] <- as.character(kcl4)

five_clusts_proj3_1 <- plot_proj(
  proj_obj = proj_obj3, 
  point_param = c(1.5, 0.2), # size, alpha, color
  plot_limits = c(-1.1, 1.1), 
  title = "a3", 
  cex = 2, 
  axis_text_size = 4,
  is_color = TRUE) + scale_color_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e')) +
  theme(legend.position = "none")

```{r five-5clusts-projections1}
scaled_data <- scale_data_manual(five_clusts[, -5])

proj_obj1[["cluster"]] <- as.character(kcl5)

five_clusts_proj1_2 <- plot_proj(
  proj_obj = proj_obj1, 
  point_param = c(1.5, 0.2), # size, alpha, color
  plot_limits = c(-1.1, 1), 
  title = "a1", 
  cex = 2, 
  axis_text_size = 4,
  is_color = TRUE) + scale_color_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e')) +
  theme(legend.position = "none")

```{r five-5clusts-projections2}
## Second projection
projection <- cbind(
  c(0.34673,-0.35774,0.66272,0.27298),
  c(0.24023,0.01170,-0.41886,0.72707))

proj_obj2[["cluster"]] <- as.character(kcl5)

five_clusts_proj2_2 <- plot_proj(
  proj_obj = proj_obj2, 
  point_param = c(1.5, 0.2), # size, alpha, color
  plot_limits = c(-1.1, 1.2), 
  title = "a2", 
  cex = 2, 
  axis_text_size = 4,
  is_color = TRUE) + scale_color_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e')) +
  theme(legend.position = "none")
```

```{r five-5clusts-projections3}
## Third projection
projection <- cbind(
  c(0.47388,-0.20984,-0.56560,0.41644),
  c(-0.51124,-0.60184,0.06463,0.36627))

proj_obj3[["cluster"]] <- as.character(kcl5)

five_clusts_proj3_2 <- plot_proj(
  proj_obj = proj_obj3, 
  point_param = c(1.5, 0.2), # size, alpha, color
  plot_limits = c(-1.1, 1.1), 
  title = "a3", 
  cex = 2, 
  axis_text_size = 4,
  is_color = TRUE) + scale_color_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e')) +
  theme(legend.position = "none")

```{r highd-proj-clust-algo-pdf, eval=knitr::is_latex_output(), fig.pos="!ht", fig.cap="Three $2\\text{-}D$ projections from $4\\text{-}D$, for the five clusters data. The helical spiral cluster is represented in dark green, the hemisphere cluster in orange, the uniform cube-shaped cluster in purple, the blunted cone cluster in pink, and the Gaussian-shaped cluster in light green.", fig.alt="Three $2\\text{-}D$ projections of five $4\\text{-}D$ synthetic clusters: helical spiral (dark green), hemisphere (orange), uniform cube (purple), cone (pink), and Gaussian (light green), showing spatial separation and cluster geometry.", fig.width=12, fig.height=8, fig.pos="!ht"}

five_clusts_proj1_1 + five_clusts_proj2_1 + five_clusts_proj3_1 +
  five_clusts_proj1_2 + five_clusts_proj2_2 + five_clusts_proj3_2 +
  plot_layout(ncol = 3) 
```

The Langevitour displays of the four- and five-cluster $k$-means solutions further clarify why these partitions do not fully recover the underlying structure. In both visualizations, the estimated clusters show noticeable overlap, with colours repeatedly blending as the tour moves—evidence that points from different true groups are assigned to the same cluster. The $4$-cluster solution shows large, diffuse regions where several true clusters are merged, while the $5$-cluster solution partially separates some structures but still misclassifies substantial portions of multiple groups. These patterns reflect the mismatch between the nonlinear shapes present in the data and the spherical decision boundaries imposed by $k$-means. Even though the cluster validity indices favour a $4–5$ cluster solution overall, the dynamic projections reveal that neither solution provides clean separation, highlighting the importance of visual diagnostics when interpreting clustering results.

### 2025-12-12 paper-cardinalR.tex
Put clustering stats in a new script (speed up the paper rendering)

\begin{figure}[!]
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-cluster-stats-1} \caption{Comparison of clustering performance metrics (within–between ratio (wb.ratio), Dunn index, Corrected Rand index, and variation of information (VI) across $k$-means, hierarchical, and model-based clustering methods.}\label{fig:fig-cluster-stats}
\end{figure}

Figure \ref{fig:fig-cluster-stats} shows a selection of cluster metrics for \(2-10\) clusters for each of the methods, \(k\)-means, hierarchical, and model-based. As is typical, the suggestion of the best solution varies between cluster statistics. Although the metrics differ in their preferences, several show consistent support for a \(4–5\) cluster solution. The Calinski--Harabasz index increases sharply up to four or five clusters before levelling off, the Dunn and Pearson gamma indices attain local maxima in this range, and both the silhouette width and within-cluster sum of squares show diminishing improvement beyond five clusters. Together, these patterns indicate that \(4–5\) clusters provide a good balance of separation and compactness, with \(k\)-means performing slightly better than the other methods across most metrics.

>>>>

\begin{figure}[!]
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/fig-cluster-stats-1} \caption{Cluster validity metrics for solutions with $2–10$ clusters obtained using $k$-means, hierarchical, and model-based clustering. Several indices consistently suggest that $4–5$ clusters provide the best balance of separation and compactness, with $k$-means performing slightly better across metrics.}\label{fig:fig-cluster-stats}
\end{figure}

Figure \ref{fig:fig-cluster-stats} shows a selection of cluster metrics for \(2-10\) clusters for each of the methods, \(k\)-means, hierarchical, and model-based. As is typical, the suggestion of the best solution varies between cluster statistics. Although the metrics differ in their preferences, several show consistent support for a \(4-5\) cluster solution. The Calinski--Harabasz index increases sharply up to four or five clusters before levelling off, the Dunn and Pearson gamma indices attain local maxima in this range, and both the silhouette width and within-cluster sum of squares show diminishing improvement beyond five clusters. Together, these patterns indicate that \(4-5\) clusters provide a good balance of separation and compactness, with \(k\)-means performing slightly better than the other methods across most metrics.

\begin{figure}[!ht]
\includegraphics[width=1\linewidth,alt={Three $2\text{-}D$ projections of five $4\text{-}D$ synthetic clusters: helical spiral (dark green), hemisphere (orange), uniform cube (purple), cone (pink), and Gaussian (light green), showing spatial separation and cluster geometry.}]{paper-cardinalR_files/figure-latex/highd-proj-clust-algo-pdf-1} \caption{Three $2\text{-}D$ projections from $4\text{-}D$, for the five clusters data. The helical spiral cluster is represented in dark green, the hemisphere cluster in orange, the uniform cube-shaped cluster in purple, the blunted cone cluster in pink, and the Gaussian-shaped cluster in light green.}\label{fig:highd-proj-clust-algo-pdf}
\end{figure}

The Langevitour displays of the four- and five-cluster \(k\)-means solutions further clarify why these partitions do not fully recover the underlying structure. In both visualizations, the estimated clusters show noticeable overlap, with colours repeatedly blending as the tour moves---evidence that points from different true groups are assigned to the same cluster. The \(4\)-cluster solution shows large, diffuse regions where several true clusters are merged, while the \(5\)-cluster solution partially separates some structures but still misclassifies substantial portions of multiple groups. These patterns reflect the mismatch between the nonlinear shapes present in the data and the spherical decision boundaries imposed by \(k\)-means. Even though the cluster validity indices favour a \(4–5\) cluster solution overall, the dynamic projections reveal that neither solution provides clean separation, highlighting the importance of visual diagnostics when interpreting clustering results.
